Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-29 _tex/paper2.tex
Built site for gh-pages

In the conceptual-population model, we assume an idealized scenario
where researchers have access to a \emph{conceptual population} of
comparative data --data representing all repeated judgments made by each
judge for every stimulus pair produced by each pair of individuals in
the population. This assumption allows us to integrate Thurstone's
theoretical principles and propose innovations to address some of the
issues raised in Section~\ref{sec-theory-issues}.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. Each structural component defines
how observed or latent variables affect the primary latent variable of
interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
particularly when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. While these topics are beyond the scope of this
  study, readers seeking a deeper understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the primary
latent variable of interest. Additionally, the branch accounts for
idiosyncratic errors: \(e_{I}\) captures unexplained variations in
students' traits unexplained by \(X_{I}\), while \(e_{IA}\) captures
variations in texts' traits unexplained by \(T_{I}\) or \(X_{IA}\).
Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
dimension \((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns
(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
length \(n_{I} \cdot n_{A}\).

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the primary latent variable of
interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures unexplained variations in judges' bias unexplained by
\(Z_{J}\), while \(e_{JK}\) captures variations in judgments unexplained
by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \((n_{J} \cdot n_{K})\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population} of comparative data, encompassing all
repeated judgments from every judge for each pair of stimuli produced by
every individual in the population. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.
The extensions to the conceptual-population model, derived from this
realistic scenario, are concisely described in the sample-comparison
model.

@@

In the sample-comparison model, we assume a more realistic scenario than
the conceptual-population model. Specifically, we assume researchers
have access to a data sample consisting of a limited number of judgments
\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\), drawn from the conceptual population. Additionally, we
assume that judges do not make \emph{all possible comparisons} allowed
by the sample, but rather only \(n_{C}\) comparisons.

@@

Where each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary
value indicating the presence or absence of data in the \(I\) and \(J\)
vectors (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{I}\) and
\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
Specifically, we define the vector \(S_{I}\) as in equation
(\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).

@@

Similarly, each element of the vectors \(S_{IA}\) and \(S_{JK}\) is a
binary value indicating the presence or absence of data in the matrices
\(IA\) and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors
\(S_{IA}\) and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and
\(n_{J} \cdot n_{K}\) elements, respectively. Specifically, we define
the vector \(S_{IA}\) as in equation (\ref{eq-mat23}) and apply a
similar definition to \(S_{JK}\).

@@

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose the sample excludes the second
student and the third judge. Additionally, assume that the second text
from each individual and the third judgment from each judge are also
missing. Under these conditions, the vectors have the following
structure:

@@

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\) vectors. In other words, researchers can only observe texts
and judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
similar sum in vector \(S_{JK}\) represents the total sampled judgments
across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every
component of \(S_{JK}\). Essentially, this vector serves as the sampled
equivalent of the \(R\) matrix defined in equation (\ref{eq-mat12}),
where the sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).

@@

Additionally, the DAG depicts the \emph{sample design} vector \(S\) as a
causal factor influencing the sample-comparison outcome \(O^{sc}_{R}\).
A square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a given variable \citep[pp.~32]{Neal_2020}. Thus, the vector
\(S\) is responsible for selecting the subset of observation from the
conceptual-population outcome \(O^{cp}_{R}\), generating the
sample-comparison outcome \(O^{sc}_{R}\).

@@

@@

Conversely, we make the scenario depicted in
Section~\ref{sec-theory-theoretical_SC1} more realistic by integrating
the comparison mechanism, that is, we assume that researchers do not
have access to \emph{all comparisons} from the sample, but rather a
sufficient number of comparisons \((n_{C})\) so that a proportion,
\(P(B>A)\), may be determined for each pair of stimuli
\citep[pp.~267]{Thurstone_1927b}. Figure~\ref{fig-cj14} illustrates this
model, incorporating both the sample and comparison mechanisms.

Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
DAG~\ref{fig-cj14_dag} depicts the sample-comparison outcome
\(O^{sc}_{R}\) as an unobserved variable, emphasizing the researchers'
inability to access such an outcome due to the comparison mechanism.
Moreover, the DAG shows the conditioned \emph{comparison design}
variable \(C\), which causally influence the observed outcome \(O_{R}\).
This variable reflects how this observed outcome is a sample drawn from
the sample-comparison outcome \(O^{sp}_{R}\). Specifically, the variable
is a vector \(C = \{ c_{1}, \dots, c_{n} \}\), where \(n\) corresponds
to the number of rows in the matrix \(R\) (refer to
Section~\ref{sec-theory-theoretical_P2}), and each \(c_{i}\) is a binary
variable indicating the presence \((1)\) or absence \((0)\) of data in
the corresponding positions of the sample-comparison outcome
\(O^{sp}_{R}\). Specifically,

>>>>

In the conceptual-population model, we assume an idealized scenario
where researchers have access to a \emph{conceptual population} of
comparative data --data representing \emph{all repeated judgments made
by each judge for every stimulus pair produced by each pair of
individuals} in the population. This assumption allows us to integrate
Thurstone's theoretical principles and propose innovations to address
some of the issues raised in Section~\ref{sec-theory-issues}.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. Each structural component defines
how observed or latent variables affect the primary latent variable of
interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
and when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the primary
latent variable of interest. Additionally, the branch accounts for
idiosyncratic errors: \(e_{I}\) captures variations in students' traits
unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in texts'
traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is an
observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
Similarly, \(X_{IA}\) is an observed matrix with dimension
\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
\(n_{I} \cdot n_{A}\).

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the primary latent variable of
interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
while \(e_{JK}\) captures variations in judgments unexplained by
\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population} of comparative data, encompassing all
repeated judgments from every judge for each pair of stimuli produced by
every individual in the population. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.
The extensions to the conceptual-population model, derived from this
realistic scenario, are described in the sample-comparison model.

@@

In the sample-comparison model, we assume a more realistic scenario than
the conceptual-population model. Specifically, we assume researchers
have access to a data sample consisting of a \emph{limited number of
judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
specific number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\)}, drawn from the conceptual population. In a subsequent
step, we assume that judges \emph{do not make all possible comparisons}
allowed by the sample, but rather only \(n_{C}\) comparisons.

@@

Where each element of \(S_{I}\) and \(S_{J}\) is a binary value
indicating the presence or absence of data elements in the \(I\) and
\(J\) vectors (see equation (\ref{eq-mat11})). Thus, the vectors
\(S_{I}\) and \(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements,
respectively. Specifically, we define the vector \(S_{I}\) as in
equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).

@@

Similarly, each element of \(S_{IA}\) and \(S_{JK}\) is a binary value
indicating the presence or absence of data rows in the matrices \(IA\)
and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{IA}\)
and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. Specifically, we define the vector \(S_{IA}\) as
in equation (\ref{eq-mat23}) and apply a similar definition to
\(S_{JK}\).

@@

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge. Additionally, they do not include the
second text from each individual and the third judgment from each judge.
Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
these vectors take the following structure:

@@

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
similar sum in vector \(S_{JK}\) represents the total sampled judgments
across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). The sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).

@@

@@

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}. Essentially,
\(S\) serves as the sample equivalent for the \(R\) matrix (see equation
(\ref{eq-mat12})), that is, this vector determines \emph{which repeated
comparisons a subset of judges makes for a subset of stimuli produced by
a subset of individuals} from the conceptual-population, ultimately
generating the sample-comparison outcome \(O^{sc}_{R}\) from the
conceptual-population outcome \(O^{cp}_{R}\).

Notably, the DAG shows that \(S\) is independent of all other variables
in the model. This implies that the conceptual model represented in
DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
judgment, judge, stimulus, and individual has the same probability of
being included in the sample as any other observation within their
respective groups \citep[pp.~16-112]{Lawson_2015}.

However, because CJ experiments rarely exhaustively pair sampled
judgments, judges, stimuli, and individuals--due to concerns about the
practical feasibility of the comparison task, like in
\citet[p.~5]{Boonen_et_al_2020}--a realistic scenario must account for
judges comparing only specific stimuli from certain individuals.

@@
